---
title: "final_project"
author: "Wanni Lei"
date: "2022-11-18"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(huxtable)
library(gtsummary)
library(tidyverse)
library(ggplot2)
library(ggpubr)
library(kableExtra)
library(gt)
library(formattable)
library(knitr)
```

```{r}
setwd("/Users/wanni/Desktop/final_project")
Bladder <- read.csv("Bladder_Cancer_Bone_Metastasis.csv")
```

## Methods
This is an observational study designed to find the potential predictors of skeletal-related event (SRE) among Bladder Cancer Patients with Bone Metastases. The associations between age at diagnosis of bladder cancer, did the patient have the symptoms at diagnosis of bone metastasis, and type of urinary diversion with SRE are what we are going to test. Dataset used in this study is manually abstracted from electronic medical records of a real clinical trial, including demographic (Sex, Ethnicity, Smoking history, et al.) and clinical characteristics (Type of urinary diversion, Did the patient have the symptoms at diagnosis of bone metastasis, Number of months from diagnosis of bone metastasis to the first skeletal-related event, et al.) of 329 patients with bone metastatic bladder cancer. This study has two cohorts, one is patients with bone metastatic bladder cancer developed an SRE within the first 2 months after diagnosis of metastatic disease, and the other is patients with bone metastatic bladder cancer did not develop an SRE within the first 2 months after diagnosis of metastatic disease. We excluded the patients who had SRE at the baseline. 

#### Why use 2 months to represent the risk of SRE
```{r}
sreBladder2m <- Bladder %>%
    filter(months_from_bony_met_dx_to_1st_sre != 0) %>% ## exclude SRE at baseline
    mutate(sreStatus = ifelse(months_from_bony_met_dx_to_1st_sre <= 2, 
                              "SRE in 2 months", "No SRE in 2 months"))

c2m <- sreBladder2m %>% filter(sreStatus=="SRE in 2 months")

prop2m <- nrow(c2m)/nrow(sreBladder2m)
prop2m %>%round(3)   



```
Because more than 60% patients developed SRE within 2 months, so we use developed SRE within 2 months to represent the risk of SRE.



#### Association between age at diagnosis of bladder cancer and SRE development.
We compared the mean age at diagnosis of bladder cancer within the two cohorts mentioned above. First we draw boxplots to get the first insight of our data.

```{r}
y2m <- sreBladder2m %>% filter(sreStatus=="SRE in 2 months") ##SRE in 2 month
n2m <- sreBladder2m %>% filter(sreStatus=="No SRE in 2 months") ## no sre in 2 month

## The data is quite symetric and the sample size is large, so we can use t test
sreBladder2m %>% 
  ggplot(aes(x = sreStatus, y = age_at_bladder_cancer_diagnosis)) +
  geom_boxplot()+
  labs(title = "Age at diagnosis of bladder cancer by two SRE groups",
       x="Wether develped SRE within 2 moth", 
       y ="Age")

var(y2m$age_at_bladder_cancer_diagnosis)
var(n2m$age_at_bladder_cancer_diagnosis) 
## the variances in two groups are quite different,
##therefore, we use false in t.test
```
The boxplots show that the data in the two groups are quite symmetric and don't have obvious outliers. Moreover, the variances in the two are quite different, so we can use the t.test with `r var.equal=FALSE` to test the mean of age in the two cohorts.

The Null hypothesis is: There is no association between age at diagnosis of bladder cancer with developed an SRE within the first 2 months. And the Alternative hypothesis is: There is an association between age at diagnosis of bladder cancer with developed an SRE within the first 2 months.
```{r}
## perform two sample t test to test the mean 
tResults2 <- t.test(age_at_bladder_cancer_diagnosis ~ sreStatus,
                    var.equal=FALSE, data=sreBladder2m) 

paste("p value is ", tResults2$p.value %>% round(4))
paste("95% CI of true difference in means is (", tResults2$conf.int[1] %>% round(4), 
      ",", tResults2$conf.int[2]%>% round(4),")" )

```

#### Association between did the patient have the symptoms at diagnosis of bone metastasis and SRE development.

The null hypothesis is: There is no association between did the patient have the symptoms at diagnosis of bone metastasis and SRE development. And the alternative hypothesis is there is an association between did the patient have the symptoms at diagnosis of bone metastasis and SRE development. First we use
chi-square's test to compute the expected value of the obtained contigency table, we found that two  of the expected counts in this table is too small (<5), so Fisher’s Exact Test will be more appropriate. Therefore, we used Fisher’s exact test to test the dependency between SRE and did the patient have the symptoms at diagnosis of bone metastasis.
```{r}
## visualize of contigency table
#table(sreBladder2m$symptomatic_at_bony_met_diagnosis,sreBladder2m$sreStatus)

## clean the data
sreBladder2m_clean <- sreBladder2m%>% 
  mutate(symptomatic_at_bony_met_diagnosis=
           ifelse(symptomatic_at_bony_met_diagnosis=="no", "No",symptomatic_at_bony_met_diagnosis))

table <- table(sreBladder2m_clean$symptomatic_at_bony_met_diagnosis,sreBladder2m_clean$sreStatus)
table ## table of cleaned data
```


```{r}
expected <- chisq.test(table)$expected
expected

result1 <- fisher.test(table, conf.level = 0.95)
result1

```

#### Association between Type of urinary diversion and SRE development.

The null hypothesis is: There is no association between Type of urinary diversion and SRE development. And the alternative hypothesis is there is an association between Type of urinary diversion and SRE development. First we use
chi-square's test to compute the expected value of the obtained contigency table, we found that none of the expected counts in this table less than 5, so Chi-square Test is appropriate for this situation. Therefore, we used Chi-square Test to test the dependency Type of urinary diversion and SRE development.
```{r}
## visualize of contigency table
table2 <- table(sreBladder2m$type_of_urinary_diversion,
                sreBladder2m$sreStatus)
table2
expected <- chisq.test(table2,correct=FALSE)$expected
#expected

result1 <- chisq.test(table2, correct=FALSE)
result1

```


#### There are tables of our variables of interests.
```{r}
## change NA to missing, F to female, and m to male
Bladder2 <- Bladder  %>% 
    mutate(smoking_history = ifelse(is.na(smoking_history),"Missing", smoking_history))%>% 
  mutate(Sex = ifelse(sex =="F", "Female","male")) %>% 
  filter(!is.na(pre_bony_met_bmi))%>% 
  mutate(Age = age_at_bladder_cancer_diagnosis)%>% 
  mutate(BMI = pre_bony_met_bmi)

## table of summary statistics for categorical 
Bladder2 %>% 
  select(Sex,smoking_history,ethnicity)%>% 
  tbl_summary(label = list(smoking_history ~"Smoking history",
    ethnicity~"Ethnicity"))%>%
  modify_caption("**Table 1. Descriptive for  different characteristics of  patients**") %>%
  modify_footnote(
    all_stat_cols() ~ "Count(proportion)"
  ) %>%
  bold_labels()



## table2 is for continuous variable
Bladder2 %>%
  select(BMI, Age) %>%
  gather("Characteristics", "Value", BMI, Age) %>%
  group_by(Characteristics) %>%
  summarise(mean = mean(Value)%>%round(3),
            median = median(Value)%>%round(3),
            sd = sd(Value)%>%round(3),
            var = var(Value)%>%round(3),
            quant_25 = quantile(Value, .25)%>%round(3),
            quant_75 = quantile(Value, .75)%>%round(3),
            min = min(Value),
            max = max(Value)) %>%
  mutate(IQR = paste0(sprintf("%.2f",quant_25)," to ",sprintf("%.2f",quant_75)),
         Range = paste0(sprintf("%.2f",min)," to ",sprintf("%.2f",max))) %>%
  select(-quant_25,-quant_75) -> dat_summaries
dat_summaries %>% kbl(caption = "Table2: Summary of age and BMI") %>%
  kable_classic(full_width = F, html_font = "Cambria")
  
  
  
  
  #knitr::kable(caption = "Table2: Summary of age and bmi")






## contingency table
formattable(table) %>% 
  kbl(caption = "Table3: SRE status by wether patient have the symptoms at diagnosis of bone") %>%
  kable_classic(full_width = F, html_font = "Cambria")

  #kable("html", escape = F, caption = "Table3: SRE status by wether patient have the symptoms at diagnosis of bone")

formattable(table2) %>% 
  kbl(caption = "Table4: SRE status by Type of urinary diversion") %>%
  kable_classic(full_width = F, html_font = "Cambria")

  #kable("html", escape = F, caption = "Table4: SRE status by Type of urinary diversion")



```












































